WNBA vs NBA Earnings In Comparable Players

### 2022-11-14 Manuscript/Manuscript.tex
Fixed long lines

Since the formulation of the Women’s National Basketball Association (WNBA) in 1996, there have been never-ending debates among fans comparing the female counterpart to the much older, more popular National Basketball Association (NBA). Women’s basketball has a rich history marred by frequent sexism, fewer opportunities, and significantly less financial support.

@@

Women’s basketball began being played in 1892 at Smith College, just one year after men’s basketball was invented \cite{Shattering_The_Glass}. Due to the gender roles and cultural norms, however, women at the time were believed to be very fragile and unable to play physical sports like basketball at anywhere near the same level as men and thus played under modified rules specially created for them \cite{WNBA_Hist}. Those familiar with basketball will know that it is usually played competitively with 5 players on each team, and each player runs the full court to play offense on one end and defend the other team on the other end. In 1892, however, women initially played “9 on 9”, with the court divided into three areas: offense, defense, and midcourt. Each woman was assigned one of these courts in which they played, such that 3 women on each team were in each section. The players could not leave their section of the court to go into the other, essentially creating three separate “3 on 3” games. The other major difference from the men’s game was that women could only dribble the ball a limited number of times and could not hold it for more than three seconds, rule changed attributed to the fear that the women players might suffer “nervous fatigue” if the game were too strenuous \cite{WNBA_Hist}.

@@

Everyone now knows that such attitudes towards women’s sports are ridiculous and sexist. Certainly there are some physical characteristic differences between men and women in sports, but for every male athlete there is a just as accomplished female one. The gradual changing of perspectives on women’s athletics is seen in the evolution of women’s basketball; while men’s basketball has largely stayed the same over the course of its almost 130 year history (key changes including the legalization of the slam dunk and the implementation of the three point line), women’s basketball has shifted slowly to eventually become almost identical. The sport was played inter-collegiately for decades beginning in the late 19th century, but did not enter the professional realm until the 1970s. The sport became officially recognized at the collegiate level by the NCAA in 1982 (a shockingly late introduction), and the Women’s Basketball League (WBL), founded in 1978 and only lasting three seasons, was the first of several failed women’s professional leagues in the late 20th century. In the meantime, the NBA experienced a huge boom in popularity largely due to the emergence of superstars Larry Bird, Magic Johnson, and Michael Jordan. The massive increase in NBA reach and revenue directly led the league creating, funding, and supporting the WNBA in 1996.

@@

There are minimal differences between today's women’s basketball and the basketball that men have been playing since inception. The WNBA and NCAAW ball is slightly smaller than that used by male players, the three point line is slightly shorter, and the game clock features four 10 minute quarters as opposed to the NBA’s 12 minute quarters, but for all intents and purposes the games are the same. The rules are otherwise identical, the athletes are similarly talented, the level of play is incredibly high, and the games are very enjoyable to watch. There remains only one massive difference between the two leagues: player salaries.

@@

The difference in player salaries is a huge point of contention in women’s basketball discourse and is often the subject of intense ridicule, and it is easy to see why: top NBA players make upwards of 40 million USD per season, not including endorsement deals and other income, while the WNBA most valuable player makes less than the average office worker in New York (an exaggeration, maybe, but Elena Delle Donne made around 120 thousand USD in her MVP year with little extra endorsement money or exposure). An NBA rookie signs one contract and has enough wealth to live comfortably off for the rest of his life, often for generations; a WNBA prospect is encouraged to continue her academic studies to ensure that she is able to get a good job after her career is over. Now, granted, the league revenue is much much less than that of the NBA, but many questions still remain that this paper hopes to address: Are the players paid fairly compared to league revenue? Are they paid fairly compared to NBA players based on basketball production? How can this change?

@@

In this paper, I hope to compare not just salaries of WNBA vs NBA players, which is something that has been done numerous times, but to go further in depth and examine salaries vs team salary caps, league revenues, games played, and statistical performances for each player. The goal is to draw conclusions on if there are certain statistics in the NBA that result in higher adjusted salaries than the WNBA and vice versa, what the correlation is between them, and determine superstar salary proportions (amongst similar questions). These are related, yet more in depth questions to the ones posed by Elle Baker.

@@

The paper will begin by discussing the datasets being analyzed and how they were cleaned and prepped for analysis using Python scripts. In the next section, statistical analysis will be performed to try and answer some of the questions posed in this introduction. Following that will be a discussion on the results and hypotheses for best paths forward. Finally, there will be a conclusions section to wrap up the paper and review the results.

>>>>

Since the formulation of the Women’s National Basketball Association (WNBA) in 1996, there have been never-ending debates 
among fans comparing the female counterpart to the much older, more popular National Basketball Association (NBA). 
Women’s basketball has a rich history marred by frequent sexism, fewer opportunities, and significantly less financial support.

@@

Women’s basketball began being played in 1892 at Smith College, just one year after men’s basketball was invented 
\cite{Shattering_The_Glass}. Due to the gender roles and cultural norms, however, women at the time were believed 
to be very fragile and unable to play physical sports like basketball at anywhere near the same level as men and 
thus played under modified rules specially created for them \cite{WNBA_Hist}. Those familiar with basketball will 
know that it is usually played competitively with 5 players on each team, and each player runs the full court to 
play offense on one end and defend the other team on the other end. In 1892, however, women initially played “9 on 9”, 
with the court divided into three areas: offense, defense, and midcourt. Each woman was assigned one of these courts 
in which they played, such that 3 women on each team were in each section. The players could not leave their section 
of the court to go into the other, essentially creating three separate “3 on 3” games. The other major difference 
from the men’s game was that women could only dribble the ball a limited number of times and could not hold it for 
more than three seconds, rule changed attributed to the fear that the women players might suffer “nervous fatigue” 
if the game were too strenuous \cite{WNBA_Hist}.

@@

Everyone now knows that such attitudes towards women’s sports are ridiculous and sexist. Certainly there are some physical 
characteristic differences between men and women in sports, but for every male athlete there is a just as accomplished 
female one. The gradual changing of perspectives on women’s athletics is seen in the evolution of women’s basketball; 
while men’s basketball has largely stayed the same over the course of its almost 130 year history (key changes including 
the legalization of the slam dunk and the implementation of the three point line), women’s basketball has shifted slowly 
to eventually become almost identical. The sport was played inter-collegiately for decades beginning in the late 19th 
century, but did not enter the professional realm until the 1970s. The sport became officially recognized at the collegiate 
level by the NCAA in 1982 (a shockingly late introduction), and the Women’s Basketball League (WBL), founded in 1978 and 
only lasting three seasons, was the first of several failed women’s professional leagues in the late 20th century. In the 
meantime, the NBA experienced a huge boom in popularity largely due to the emergence of superstars Larry Bird, 
Magic Johnson, and Michael Jordan. The massive increase in NBA reach and revenue directly led the league creating, 
funding, and supporting the WNBA in 1996.

@@

There are minimal differences between today's women’s basketball and the basketball that men have been playing since 
inception. The WNBA and NCAAW ball is slightly smaller than that used by male players, the three point line is slightly 
shorter, and the game clock features four 10 minute quarters as opposed to the NBA’s 12 minute quarters, but for all 
intents and purposes the games are the same. The rules are otherwise identical, the athletes are similarly talented, 
the level of play is incredibly high, and the games are very enjoyable to watch. There remains only one massive 
difference between the two leagues: player salaries.

@@

The difference in player salaries is a huge point of contention in women’s basketball discourse and is often the subject 
of intense ridicule, and it is easy to see why: top NBA players make upwards of 40 million USD per season, not including 
endorsement deals and other income, while the WNBA most valuable player makes less than the average office worker in 
New York (an exaggeration, maybe, but Elena Delle Donne made around 120 thousand USD in her MVP year with little extra 
endorsement money or exposure). An NBA rookie signs one contract and has enough wealth to live comfortably off for 
the rest of his life, often for generations; a WNBA prospect is encouraged to continue her academic studies to 
ensure that she is able to get a good job after her career is over. Now, granted, the league revenue is much 
much less than that of the NBA, but many questions still remain that this paper hopes to address: Are the players 
paid fairly compared to league revenue? Are they paid fairly compared to NBA players based on basketball production? 
How can this change?

@@

In this paper, I hope to compare not just salaries of WNBA vs NBA players, which is something that has been done 
numerous times, but to go further in depth and examine salaries vs team salary caps, league revenues, games played, 
and statistical performances for each player. The goal is to draw conclusions on if there are certain statistics in 
the NBA that result in higher adjusted salaries than the WNBA and vice versa, what the correlation is between them, 
and determine superstar salary proportions (amongst similar questions). These are related, yet more in depth questions 
to the ones posed by Elle Baker.

@@

The paper will begin by discussing the datasets being analyzed and how they were cleaned and prepped for analysis using 
Python scripts. In the next section, statistical analysis will be performed to try and answer some of the questions 
posed in this introduction. Following that will be a discussion on the results and hypotheses for best paths forward. 
Finally, there will be a conclusions section to wrap up the paper and review the results.
